---
title: "Analysis"
output: word_document
---

```{r}
#'This script runs the statistical analysis for a given dataset.
#' 
#'Input: A csv file with variants that were either detected or not detected
#'by each caller.
#' 
#'Output: A word document with the results of the fatalistically analyses.
#'
#'Author: Georgios Karakatsoulis
#' 
#'
```



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
knitr::opts_chunk$set(results = 'asis')
```


```{r libraries, include=FALSE}
library(dplyr)
library(rstatix)
library(ggplot2)
library(pROC)
library(cutpointr)
library(glmnet)
library(caret)
```

```{r read_file, include=FALSE}

myfile_wd = gsub('Rscripts', 'Desktop/', getwd())

df = read.csv(paste0(myfile_wd, '1000_100_10_All_Freebayes_read_pos_bins_report.csv'))

colnames(df)[6] = 'sequences'

df$ref_alt = paste(df$REF, df$ALT)


df$outcome = ifelse(df$Found == 'Yes', 1, 0)

```

# Correlation between Found and the categorical variables

```{r categorical}


tmp = df %>%
  
  select(REF, ALT, bin, Found) %>%
  
  gather(., key = 'RiskFactor', value = 'RF_Value', -Found) %>%
  
  freq_table(RiskFactor, RF_Value, Found) %>%
  
  mutate(
    
    result = paste0(n, ' (', prop, '%', ')')
    
  ) %>%
  
  select(-n, -prop) %>%
  
  spread(., key = 'Found', value = 'result')

tmp$p_value = NA

tmp$p_value[which(tmp$RiskFactor == 'REF')] = chisq.test(df$Found, df$REF)$p.value

tmp$p_value[which(tmp$RiskFactor == 'ALT')] = chisq.test(df$Found, df$ALT)$p.value

a = glm(ifelse(Found == 'Yes', 1, 0) ~ bin, data = df, family = 'binomial') %>%
  
  summary() %>%
  
  coefficients()

tmp$p_value[which(tmp$RiskFactor == 'bin')] = a[2,4] 

tmp$p_value = ifelse(
  
  tmp$p_value < 0.001,
  '<0.001',
  as.character(round(tmp$p_value, 3))
)

tmp$p_value[which(duplicated(tmp$RiskFactor))] = ''

knitr::kable(tmp)


```


# Correlation between Found and the numeric variables


```{r numeric}

df_long = df %>%
  
  select(pos_chromosomal, pos_read, sequences, Found) %>%
  
  gather(., key = 'RiskFactor', value = 'RF_Value', -Found)


ggplot(data = df_long, 
       
       aes(
         x = as.factor(Found),
         fill = as.factor(Found),
         y = RF_Value)) +
  
  geom_boxplot() +
  
  labs(
    
    fill = 'Found',
    x = 'Found') +
  
  facet_wrap(~RiskFactor, scales = 'free_y') 



tmp = df %>%
  
  select(pos_chromosomal, pos_read, sequences, Found) %>%
  
  group_by(Found) %>%
  
  get_summary_stats() %>%

  select(-iqr, -mad, -se, -ci, -min, -max) %>%
    
  arrange(variable) %>%
  
  rename('RiskFactor' = 'variable')

tmp$p_value = NA
  
a = glm(ifelse(Found == 'Yes', 1, 0) ~ pos_chromosomal, data = df, family = 'binomial') %>%
  
  summary() %>%
  
  coefficients()

tmp$p_value[which(tmp$RiskFactor == 'pos_chromosomal')] = a[2,4] 

a = glm(ifelse(Found == 'Yes', 1, 0) ~ pos_read, data = df, family = 'binomial') %>%
  
  summary() %>%
  
  coefficients()

tmp$p_value[which(tmp$RiskFactor == 'pos_read')] = a[2,4] 


a = glm(ifelse(Found == 'Yes', 1, 0) ~ sequences, data = df, family = 'binomial') %>%
  
  summary() %>%
  
  coefficients()

tmp$p_value[which(tmp$RiskFactor == 'sequences')] = a[2,4] 


tmp$p_value = ifelse(
  
  tmp$p_value < 0.001,
  '<0.001',
  as.character(round(tmp$p_value, 3))
)

tmp$p_value[which(duplicated(tmp$RiskFactor))] = ''

knitr::kable(tmp)

df_scale = df %>%
  
  select(outcome, pos_chromosomal, sequences, pos_read) %>%
  
  mutate_at(c('pos_chromosomal', 'pos_read', 'sequences'), scale)

tmp = matrix(nrow = 3, ncol = 5) %>%
  
  as.data.frame()

colnames(tmp) = c('RiskFactor', 'optimal_cutpoint', 'acc', 'sensitivity', 'specificity')

cutpointr(df_scale,
          x = pos_chromosomal,
          class = outcome,
          pos_class = 1,
          neg_class = 0,
          method = maximize_metric,
          metric = youden) -> cp

a = summary(cp)

tmp[1, ]= a$cutpointr %>%
  
  as.data.frame() %>%
  
  mutate(RiskFactor = 'pos_chromosomal') %>%
  
  select(all_of(colnames(tmp)))


cat('Pos_Chromosomal')

plot(cp)


cutpointr(df_scale,
          x = pos_read,
          class = outcome,
          pos_class = 1,
          neg_class = 0,
          method = maximize_metric,
          metric = youden) -> cp

a = summary(cp)

tmp[2, ]= a$cutpointr %>%
  
  as.data.frame() %>%
  
  mutate(RiskFactor = 'pos_read') %>%
  
  select(all_of(colnames(tmp)))

cat('Pos_read')

plot(cp)


cutpointr(df_scale,
          x = sequences,
          class = outcome,
          pos_class = 1,
          neg_class = 0,
          method = maximize_metric,
          metric = youden) -> cp

a = summary(cp)

tmp[3, ]= a$cutpointr %>%
  
  as.data.frame() %>%
  
  mutate(RiskFactor = 'sequences') %>%
  
  select(all_of(colnames(tmp)))

cat('Number of Sequences')

plot(cp)

knitr::kable(tmp)


```


# Correlation between Found, REF and ALT

```{r ref_alt}

tmp = df %>%
  
  freq_table(REF, ALT, Found) %>%
  
  filter(Found == 'Yes') %>%
  
  select(-Found) %>%
  
  tidyr::complete(., REF, ALT, fill = list(n = 0, prop = 0)) %>%
  
  filter(REF != ALT) %>%
  
  rename(
    
    'N_found' = 'n',
    'Pct_found' = 'prop'
    
  )
  
  
knitr::kable(tmp)


```

# Correlation of Position Chromosomal with Found (within REF and ALT)

```{r boxplot_pos_chromosomal}

df$REF = paste('REF:', df$REF)

ggplot(data = df, 
       
       aes(
         x = ALT,
         fill = as.factor(Found),
         y = pos_chromosomal)) +
  
  geom_boxplot() +
  
  labs(fill = 'Found') +
  
  facet_wrap(~REF)

```


# Correlation of Position Read with Found (within REF and ALT)

```{r boxplot_pos_read}

ggplot(data = df, 
       
       aes(
         x = ALT,
         fill = as.factor(Found),
         y = pos_read)) +
  
  geom_boxplot() +
  
  labs(fill = 'Found') +
  
  facet_wrap(~REF)

```



# Correlation of N of sequences Read with Found (within REF and ALT)

```{r boxplot_n_of_seq}

ggplot(data = df, 
       
       aes(
         x = ALT,
         fill = as.factor(Found),
         y = sequences)) +
  
  geom_boxplot() +
  
  labs(fill = 'Found', y = 'N of sequences') +
  
  facet_wrap(~REF)

```


```{r data_for_prediction}

df$REF = gsub('REF: ', '', df$REF)

df$Found = ifelse(df$Found == 'Yes', 1, 0)

rfs = c('pos_chromosomal', 'pos_read', 'REF', 'ALT', 'sequences')


```

# Logistic Regression Model (with interaction terms)

```{r logit_model, eval = T}

df1 = df %>%
  
  mutate(
    
    pos_chromosomal = scale(pos_chromosomal)
    
    )

model = glm(Found ~ pos_chromosomal * REF + pos_chromosomal * ALT +
              
              pos_read * REF + pos_read * ALT,
            family = 'binomial',
            data = df1)


tmp2 = summary(model)$coefficients %>%
  
  as.data.frame() %>%
  
  transmute(
    
    Riskfactor = rownames(.),
    OR = round(exp(Estimate),2),
    
    CI = paste0(
      
      '(',
      
      round(exp(Estimate - 1.96 * `Std. Error`), 2),
      ', ',
      round(exp(Estimate + 1.96 * `Std. Error`), 2)
      
      ,
      ')'
      
      
    ),
    p = round(`Pr(>|z|)`,3)
    
    ) %>%
  
  mutate(p = ifelse(p == 0, '<0.001', as.character(p))) %>%
  
  filter(Riskfactor != '(Intercept)') 

rownames(tmp2) = NULL

# tmp2$Riskfactor = gsub('scale\\(pos_chromosomal)', 'pos_chromosomal', tmp2$Riskfactor)

knitr::kable(tmp2)


myroc = pROC::roc(
  
  df1$Found,
  model$fitted.values,
  plot      = F,
  print.auc = T,
  ci        = T
  
)


ggroc(
  
  myroc,
  colour      = 'steelblue',
  size        = 1,
  legacy.axes = T
  
) +
  
  geom_segment(
    
    aes(x = 0, y = 0, xend = 1, yend = 1),
    
    linetype = 'dashed',
    
    colour   = 'darkgrey'
  ) +
  
  theme_minimal() +
  
  geom_segment(
    
    aes(x = 0, y = 1, xend = 1, yend = 1),
    
    linetype = 'dashed'
    
  ) +
  
  geom_segment(
    
    aes(x = 0, y = 0, xend = 0, yend = 1),
    
    linetype = 'dashed'
    
  ) +
  
  geom_segment(
    
    aes(x = 1, y = 1, xend = 1, yend = 0),
    
    linetype = 'dashed'
    
  ) +
  
  geom_segment(
    
    aes(x = 0, y = 0, xend = 1, yend = 0),
    
    linetype = 'dashed'
    
  ) +
  
  geom_text(
    
    aes(
      
      x     = 0.75,
      y     = 0.45,
      label = paste0(
        
        'AUC = ',
        round(myroc$ci[2], 3),
        ' (',
        round(myroc$ci[1], 3),
        '-',
        round(myroc$ci[3], 3),
        ')'
        
      )
      
    ),
    
    size = 4
    
  )


```


# Logistic Regression Model (without interaction terms)

```{r logit_model_no_interaction}

rfs = c('pos_chromosomal', 'pos_read', 'REF', 'ALT', 'sequences')

df1 = df %>%
  
  mutate(
    
    pos_chromosomal = scale(pos_chromosomal),
    pos_read = scale(pos_read),
    sequences = scale(sequences)
    
    )

model = glm(Found ~ pos_chromosomal + REF + ALT + pos_read,
            family = 'binomial',
            data = df1)


tmp2 = summary(model)$coefficients %>%
  
  as.data.frame() %>%
  
  transmute(
    
    Riskfactor = rownames(.),
    OR = round(exp(Estimate),2),
    
    CI = paste0(
      
      '(',
      
      round(exp(Estimate - 1.96 * `Std. Error`), 2),
      ', ',
      round(exp(Estimate + 1.96 * `Std. Error`), 2)
      
      ,
      ')'
      
      
    ),
    p = round(`Pr(>|z|)`,3)
    
    ) %>%
  
  mutate(p = ifelse(p == 0, '<0.001', as.character(p))) %>%
  
  filter(Riskfactor != '(Intercept)') 

rownames(tmp2) = NULL

# tmp2$Riskfactor = gsub('scale\\(pos_chromosomal)', 'pos_chromosomal', tmp2$Riskfactor)

knitr::kable(tmp2)


myroc = pROC::roc(
  
  df1$Found,
  model$fitted.values,
  plot      = F,
  print.auc = T,
  ci        = T
  
)


ggroc(
  
  myroc,
  colour      = 'steelblue',
  size        = 1,
  legacy.axes = T
  
) +
  
  geom_segment(
    
    aes(x = 0, y = 0, xend = 1, yend = 1),
    
    linetype = 'dashed',
    
    colour   = 'darkgrey'
  ) +
  
  theme_minimal() +
  
  geom_segment(
    
    aes(x = 0, y = 1, xend = 1, yend = 1),
    
    linetype = 'dashed'
    
  ) +
  
  geom_segment(
    
    aes(x = 0, y = 0, xend = 0, yend = 1),
    
    linetype = 'dashed'
    
  ) +
  
  geom_segment(
    
    aes(x = 1, y = 1, xend = 1, yend = 0),
    
    linetype = 'dashed'
    
  ) +
  
  geom_segment(
    
    aes(x = 0, y = 0, xend = 1, yend = 0),
    
    linetype = 'dashed'
    
  ) +
  
  geom_text(
    
    aes(
      
      x     = 0.75,
      y     = 0.45,
      label = paste0(
        
        'AUC = ',
        round(myroc$ci[2], 3),
        ' (',
        round(myroc$ci[1], 3),
        '-',
        round(myroc$ci[3], 3),
        ')'
        
      )
      
    ),
    
    size = 4
    
  )


```




# LASSO penalty in logistic regression (with interaction terms)


```{r lasso, eval = T}

rfs = c('pos_chromosomal', 'pos_read', 'REF', 'ALT', 'sequences')


df1 = df %>%
  
  select(Found, all_of(rfs)) %>%
  
  na.omit() %>%
  
  mutate(
    
    pos_chromosomal = scale(pos_chromosomal),
    pos_read = scale(pos_read)
    
    )


set.seed(123)

training.samples <- df1$Found %>% 
  createDataPartition(p = 0.8, list = FALSE)

train.data  <- df1[training.samples, ]

test.data <- df1[-training.samples, ]

# Dummy code categorical predictor variables
x <- model.matrix(Found~ pos_chromosomal * REF + pos_chromosomal * ALT + 
                    
                    pos_read * REF + pos_read * ALT + 
                    
                    sequences * REF + sequences * ALT, train.data)[,-1]

# Convert the outcome (class) to a numerical variable
y <- train.data$Found

# Find the best lambda using cross-validation
set.seed(123) 

cv.lasso <- cv.glmnet(x, y, alpha = 1, family = "binomial")

# Fit the final model on the training data
model <- glmnet(x, y, alpha = 1, family = "binomial",
                lambda = cv.lasso$lambda.min)

# Display regression coefficients
tmp2 = exp(coef(model))

tmp2 = data.frame(
  
  rownames(tmp2),
  round(as.vector(tmp2),2)
  
)

colnames(tmp2) = c('Riskfactor', 'OR')

tmp2 = tmp2 %>%
  
  filter(Riskfactor != '(Intercept)')

knitr::kable(tmp2)


# Make predictions on the test data
x.test <- model.matrix(Found ~ pos_chromosomal * REF + pos_chromosomal * ALT + 
                    
                    pos_read * REF + pos_read * ALT + 
                    
                    sequences * REF + sequences * ALT, test.data)[,-1]

probabilities <- model %>%
  
  predict(newx = x.test)

myroc = pROC::roc(
  
  test.data$Found,
  probabilities,
  plot      = F,
  print.auc = T,
  ci        = T
  
)


ggroc(
  
  myroc,
  colour      = 'steelblue',
  size        = 1,
  legacy.axes = T
  
) +
  
  geom_segment(
    
    aes(x = 0, y = 0, xend = 1, yend = 1),
    
    linetype = 'dashed',
    
    colour   = 'darkgrey'
  ) +
  
  theme_minimal() +
  
  geom_segment(
    
    aes(x = 0, y = 1, xend = 1, yend = 1),
    
    linetype = 'dashed'
    
  ) +
  
  geom_segment(
    
    aes(x = 0, y = 0, xend = 0, yend = 1),
    
    linetype = 'dashed'
    
  ) +
  
  geom_segment(
    
    aes(x = 1, y = 1, xend = 1, yend = 0),
    
    linetype = 'dashed'
    
  ) +
  
  geom_segment(
    
    aes(x = 0, y = 0, xend = 1, yend = 0),
    
    linetype = 'dashed'
    
  ) +
  
  geom_text(
    
    aes(
      
      x     = 0.75,
      y     = 0.45,
      label = paste0(
        
        'AUC = ',
        round(myroc$ci[2], 3),
        ' (',
        round(myroc$ci[1], 3),
        '-',
        round(myroc$ci[3], 3),
        ')'
        
      )
      
    ),
    
    size = 4
    
  )


```


# LASSO penalty in logistic regression (without interaction terms)

```{r lasso_no_interaction}

rfs = c('pos_chromosomal', 'pos_read', 'REF', 'ALT', 'sequences')


df1 = df %>%
  
  select(Found, all_of(rfs)) %>%
  
  na.omit() %>%
  
  mutate(
    
    pos_chromosomal = scale(pos_chromosomal),
    pos_read = scale(pos_read)
    
    )


set.seed(123)

training.samples <- df1$Found %>% 
  createDataPartition(p = 0.8, list = FALSE)

train.data  <- df1[training.samples, ]

test.data <- df1[-training.samples, ]

# Dummy code categorical predictor variables
x <- model.matrix(
  
  Found~ pos_chromosomal + REF + ALT + pos_read + sequences,
  train.data)[,-1]

# Convert the outcome (class) to a numerical variable
y <- train.data$Found

# Find the best lambda using cross-validation
set.seed(123) 

cv.lasso <- cv.glmnet(x, y, alpha = 1, family = "binomial")

# Fit the final model on the training data
model <- glmnet(x, y, alpha = 1, family = "binomial",
                lambda = cv.lasso$lambda.min)

# Display regression coefficients
tmp2 = exp(coef(model))

tmp2 = data.frame(
  
  rownames(tmp2),
  round(as.vector(tmp2),2)
  
)

colnames(tmp2) = c('Riskfactor', 'OR')

tmp2 = tmp2 %>%
  
  filter(Riskfactor != '(Intercept)')

knitr::kable(tmp2)


# Make predictions on the test data
x.test <- model.matrix(Found ~ pos_chromosomal + REF + ALT + 
                    
                    pos_read + sequences, test.data)[,-1]

probabilities <- model %>%
  
  predict(newx = x.test)

myroc = pROC::roc(
  
  test.data$Found,
  probabilities,
  plot      = F,
  print.auc = T,
  ci        = T
  
)


ggroc(
  
  myroc,
  colour      = 'steelblue',
  size        = 1,
  legacy.axes = T
  
) +
  
  geom_segment(
    
    aes(x = 0, y = 0, xend = 1, yend = 1),
    
    linetype = 'dashed',
    
    colour   = 'darkgrey'
  ) +
  
  theme_minimal() +
  
  geom_segment(
    
    aes(x = 0, y = 1, xend = 1, yend = 1),
    
    linetype = 'dashed'
    
  ) +
  
  geom_segment(
    
    aes(x = 0, y = 0, xend = 0, yend = 1),
    
    linetype = 'dashed'
    
  ) +
  
  geom_segment(
    
    aes(x = 1, y = 1, xend = 1, yend = 0),
    
    linetype = 'dashed'
    
  ) +
  
  geom_segment(
    
    aes(x = 0, y = 0, xend = 1, yend = 0),
    
    linetype = 'dashed'
    
  ) +
  
  geom_text(
    
    aes(
      
      x     = 0.75,
      y     = 0.45,
      label = paste0(
        
        'AUC = ',
        round(myroc$ci[2], 3),
        ' (',
        round(myroc$ci[1], 3),
        '-',
        round(myroc$ci[3], 3),
        ')'
        
      )
      
    ),
    
    size = 4
    
  )


```
